Gene expression profiling of reproductive meristem identities in rice inflorescence by laser microdissection and RNA sequencing.

### 2016-01-27 ms.Rmd
Round 2 revisions Fix clade labelling for new version of figtree Use oryzr to annotate cluster genes

---
title: Gene expression profiling of reproductive meristem identities in rice inflorescence
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
bibliography: bib/papersLib.bibtex
---

@@

In rice, inflorescence architecture is established at early stages of reproductive development and contributes directly to grain yield potential. After induction of flowering, the complexity of branching and therefore the number of seeds on the panicle are determined by the activity of different meristem types and the timing of transitions between them. Although some of the genes involved in these transitions have been identified, understanding of the network of transcriptional regulators controlling this process is lacking. To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp. *japonica* cv. Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types. A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed, and a number of candidate genes were identified that may be involved in branching. This new resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.

@@

The branched inflorescence of rice is a compound raceme, classified as a panicle, composed of a rachis (the main axis), primary branches, higher order branches, and spikelets. Grain yield is determined by the complexity and arrangement of branches and spikelets, which is specified in the early stages of panicle development by the establishement and activity of apical and axillary meristems in two phases. The timing of rachis meristem abortion specifies the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems conditions the complexity of branching and thereby the number of seeds on the panicle. After spikelet differentiation, branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk]. The landscape of gene expression is a characteristic that differentiates meristem types and is involved in the control of meristem identity transitions. Mutant analyses and mapping of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq]. Some of these genes are involved in the patterning of axillary meristem and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq]. A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].

Understanding the events during development that determine panicle characteristics, such as branching complexity and its plasticity, will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the differences in gene expression between different meristem types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *Oryza sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations. After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches. During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelets. The PBM and secondary branch meristem (SBM) are both converted to a terminal spikelet meristem (SM). Terminal and axillary SMs develop into floret meristems (FM) and produce a single floret [@Ikeda:2004tk]. A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD). After differentiation of the flag leaf primordium the first bract primordium appeared opposite to the flag leaf, and the first stage of RM appeared at the base of the bract primordium (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a). After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b). After the formation of primary branches, the rachis meristem stopped growing, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c). Each PBM produced a number of AMs as they elongated. For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d). The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million uniquely mapped reads within genes, after excluding reads likely to have originated from rRNA and tRNA. Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage (Data S1). As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with patterns that had previously been reported. `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`). The expression of a further four uncharacterized genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization. Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure ", figCount, "f_inSitu"))`). These results suggest that the tissues used for RNA sequencing were accurately dissected and represent the intended meristem types.

@@

Other transcriptomic studies using whole inflorescences [*e.g.* @Furutani:2006hx], including inflorescences at later developmental stages than described here [*e.g.* @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im], have highlighted the importance of transcription factors in rice panicle development. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections [@Kobayashi:2012fi]. This microarray study identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in plants with defects in inflorescence development [@Kobayashi:2012fi]. All four were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems. In contrast to these studies, only the meristematic regions of young panicles were collected by LMD in the RNAseq dataset described in this article, resulting in a narrow-scale transcriptomic analysis describing expression in these tissues alone.

@@

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM to SM). Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], and excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters (*ALS1* and *MDR12*), while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. Transgenic rice plants presumably expressing *GH3-8* in the endogenous pattern of *FLO-LFY HOMOLOG OF RICE* (*RFL*) [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co] have smaller panicles with fewer branches [@Yadav:2011db], and GH3-8 conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type. Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues. Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM both support significant changes in gene expression between apical and axillary meristems.

Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) is not highly expressed (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). *MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* was not reported [@Yoshida:2013ff]. Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. *TAW1* was recovered from the same cluster as *MADS55*. The presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components. Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset. Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), whilst the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one affected by the RNAi construct used against *TAW1*. Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). This transition corresponds to the acquisition of determinate fate by the axillary meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity. *AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy]. Stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di]. Comparison of these datasets with the rice LMD dataset presented here suggests some differences in expression of MIKC^C^-type MADS genes between these species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx]. *SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx]. The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). In *A. thaliana*, *miR319a* targets *TEOSINTE BRANCHED*/*CYCLOIDEA*/*PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc]. No direct interaction with *SPL* genes or role in reproductive development has been reported for *mrR319a* in rice. Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf]. WFP also binds to the promoters of genes that are involved in the determination of inflorescence architecture, such as *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1*, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

@@

This report describes the changes in gene expression that occur during the early stages of rice panicle development. Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM. The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem. These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.

@@

**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**. Geneset enrichment analysis (GSEA) of transcription factor and other transcriptional regulator families. Test statistics were calculated from the L~2~FC values between the stage of interest and all other stages for the expressed genes in each geneset. The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.

>>>>

---
title: Gene expression profiling of reproductive meristem identities in rice inflorescence by laser microdissection and RNA sequencing.
csl: csl/the-plant-journal.csl
header-includes:
  \usepackage{graphicx}
  \usepackage[utf8]{inputenc}
  \DeclareUnicodeCharacter{2265}{$\geq$}
output:
  word_document:
    fig_caption: yes
    reference_docx: csl/wordStyles.docx
  BiocStyle::html_document:
    dev: CairoPDF
    fig_caption: yes
    force_captions: yes
    highlight: pygments
    number_sections: yes
bibliography: bib/papersLib.bibtex
---

@@

In rice, inflorescence architecture is established at early stages of reproductive development and contributes directly to grain yield potential. After induction of flowering, the complexity of branching and therefore the number of seeds on the panicle are determined by the activity of different meristem types and the timing of transitions between them. Although some of the genes involved in these transitions have been identified, understanding of the network of transcriptional regulators controlling this process is lacking. To address this, we used a precise laser microdissection and RNA sequencing approach in *Oryza sativa* ssp. *japonica* cv. Nipponbare to produce quantitative data that describe the landscape of gene expression in four different meristem types: the rachis meristem, the primary branch meristem, the elongating primary branch meristem (including axillary meristems) and the spikelet meristem. A switch in expression profile between apical and axillary meristem types followed by more gradual changes during transitions in axillary meristem identity was observed, and a number of genes potentially involved in branching were identified. This resource will be vital for a mechanistic understanding of the link between inflorescence development and grain yield.

@@

The branched inflorescence of rice is a compound raceme, classified as a panicle, and is composed of a rachis (the main axis), primary branches, higher order branches, and spikelets. The number of seeds on the panicle is influenced by the complexity and arrangement of branches and spikelets. The establishement and activity of apical and axillary meristems conditions branching in two phases during the early stages of panicle development: the timing of rachis meristem abortion determines the number of primary branches, whilst the transition of indeterminate branch meristems to determinate spikelet meristems specifies the complexity of branching. After spikelet differentiation, branching complexity is fixed and the rachis and branches elongate rapidly and heading and flowering occurs [@Ikeda:2004tk]. The landscape of gene expression is a characteristic that differentiates meristem types and is involved in the control of meristem identity transitions. Mutant analyses and mapping of quantitative trait loci have identified a number of genes required for the initiation and development of panicles, as well as genes controlling the number and size of grains and panicles [@Xing:2010de; @Wang:2011kq]. Some of these genes are involved in the patterning of axillary meristems and panicle branching, such as the nuclear regulatory factor-encoding genes *MONOCULM 1*, *LAX PANICLE 1* and *LAX PANICLE 2* [@Wang:2011kq]. A large set of genes related to floral development, which may also affect panicle architecture, has been identified [@Yoshida:2011bo], some of which may have been under selection during domestication or are associated with crop improvement in Asian rice  [@Xing:2010de; @He:2011io; @Wang:2011kq; @Xu:2012hq; @Ikeda:2013go].

Understanding the events during development that determine panicle characteristics such as branching complexity and its plasticity will be vital for sustainable improvement of rice yield potential using targeted breeding programs. Despite advances in the characterization of individual genes and their interactions, a complete understanding of the control of panicle morphology and grain yield will require mechanistic studies explaining the interactions of gene regulatory network components with each other and with the environment [@Azpeitia:2013jia]. One step towards this goal is to describe the differences in gene expression between different meristem types during development. In this article, we describe the measurement and analysis of genome-wide expression in meristematic tissues from the early stages of panicle development in *Oryza sativa* ssp. *japonica* cv. Nipponbare, using a precise laser microdissection and RNA sequencing approach.

@@

In rice, the start of the reproductive phase and subsequent initiation of inflorescence development involves fast morphological transformations. After the establishment of the rachis meristem (RM), some cells differentiate into primary branch meristem (PBM) for the formation and elongation of primary branches. During primary branch elongation, the PBM can give rise to axillary meristems (AM), which may differentiate into secondary and higher-order branches or be converted directly to spikelet meristem (SM). The PBM and secondary branch meristem (SBM) are both converted to a terminal SM. Each terminal or axillary SM produces one floret meristem (FM), which differentiates into a single floret [@Ikeda:2004tk]. A detailed histological analysis of rice inflorescences was performed to select four morphologically distinct inflorescence meristem types for laser microdissection (LMD). After differentiation of the flag leaf, the first stage of reproductive development is the conversion of the shoot apical meristem (SAM) to the RM. The first bract primordium was produced opposite the flag leaf (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`a). After the first stage of reproductive growth the rachis meristem produced PBMs in the axils of newly developed bracts (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`b). After the formation of primary branches, RM activity was lost, bract growth ceased and primary branches elongated (ePBM) (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`c). Each PBM produced a number of AMs as they elongated. For this analysis, SM differentiation was considered as the final stage of panicle development (`r I(pasteLabel("Figure ", figCount, "f_lmdFigure"))`d). The RM, PBM and ePBM/AM stages are the indeterminate stages in which meristematic cells are maintained, whilst the SM has a determinate fate in which the stem cell activity will be lost and from which florets will differentiate [@Ikeda:2004tk].

@@

At least `r round(min(st_libStats$"Reads (M)"), 0)` million single-end, 50-base reads were produced from each library, yielding between `r round(min(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` and `r round(max(st_libStats$"Reads in genes (M)", na.rm = TRUE), 0)` million uniquely mapped reads within genes, after excluding reads likely to have originated from rRNA and tRNA. Using strict cut-offs for gene expression, `r length(unique(readRDS("output/expressedGenes/expressedGenesAll.Rds")))` unique genes were detected in two or more libraries from at least one stage (Data S1). As a preliminary assessment of the RNA sequencing (RNAseq) results, the qualitative and quantitative expression of `r nIS` genes were compared with previously reported patterns. `r nPos` of the `r nIS` genes were detected in two or more libraries from at least one meristem type in the LMD dataset (`r I(pasteLabel("Figure S", s_figCount, "sf_isGenesTpm"))` and `r I(pasteLabel("Table S", s_tableCount, "st_refTable"))`). The expression of a further four uncharacterized genes that were detected in specific meristem types by RNAseq was confirmed by RNA *in situ* hybridization. Each of the four genes was detected in the same meristem types by both methods (`r I(pasteLabel("Figure ", figCount, "f_inSitu"))`). These results suggest that the tissues used for RNA sequencing were accurately dissected and represent the intended meristem types.

@@

Several previous transcriptomic studies have measured gene expression in whole rice inflorescences at later stages of development than those described here [*e.g.* @Wang:2010im; @Sato:2011eh; @Sharma:2012kq; @Khanday:2013bz; @Jiang:2014im]. Another study used microarray experiments at earlier stages to highlight the importance of transcription factors in panicle development [@Furutani:2006hx]. LMD has also been used for precise control of developmental stage during the collection of whole panicle sections for microarrays [@Kobayashi:2012fi]. This identified three MADS genes, which are members of the *SQUA*-like (*FUL*-like) clade, that are co-expressed with *PANICLE PHYTOMER2* (*PAP2*), and a quadruple knockdown of these four genes resulted in defects in inflorescence development [@Kobayashi:2012fi]. All four genes were detected in all stages in the RNAseq datset (Data S1), confirming their expression in reproductive meristems. In contrast to previous transcriptomic studies, only the meristematic regions of young panicles were collected by LMD for the RNAseq dataset presented in this article to restrict the measurement of gene expression to those tissues.

@@

Clusters `r getClusterName('dec', 'small')` and `r getClusterName('inc', 'small')` contain genes that change in expression between apical (RM) and axillary meristems (PBM to SM). Axillary meristem initiation requires auxin synthesis and transport in *Arabidopsis thaliana* and maize [reviewed by @Gallavotti:2013ki], and excess auxin may be inactivated by conjugation to amino acids by GH3 enzymes [@Staswick:2005cv], but these processes are not as well understood in rice. Cluster `r getClusterName('inc', 'small')` contains the auxin response factor *ARF6A* and the small auxin-up RNA *SAUR33*, as well as two ABC transporters (*ALS1* and *MDR12*), while the complementary cluster `r getClusterName('dec', 'small')` contains four uncharacterized genes with annotations relating to auxin (*IAA10* and *IAA16*, two auxin-responsive transcriptional regulators; *ABCB27*, an ABC transporter; and *LOC_Os01g63770*, an AUX/LAX transporter), and the GH3 gene related to auxin homeostasis, *GH3-8*. Transgenic rice plants presumably expressing *GH3-8* in the endogenous pattern of *FLO-LFY HOMOLOG OF RICE* (*RFL*) [*i.e.* PBM and SBM; @IkedaKawakatsu:2012co] have smaller panicles with fewer branches [@Yadav:2011db], and GH3-8 conjugates amino acids to auxin *in vitro* [@Ding:2008hv], suggesting that the panicle phenotype may be related to reduced auxin signalling. Although auxin response factors and Aux/IAA response genes are generally more strongly expressed at earlier stages (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`), the presence of auxin-related genes in complementary clusters suggests that the transcriptional effects of auxin signalling are dependent on meristem type. Understanding of the relationship between auxin and meristem specification will require functional analysis of the components of the regulatory network involved in the response in these tissues. Nevertheless, the difference in expression of annotated transcription factor genes between RM and PBM samples and the recovery of co-expression clusters of primarily uncharacterized genes that appear to be either switched on or off between the RM and the PBM indicate a significant change in gene expression between apical and axillary meristems.

Two *STMADS11*-like (*SVP*-like) MADS genes, *MADS22* and *MADS55*, decrease in expression between PBM and SM, whilst the third (*MADS47*) was not detected (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`a). *MADS22* and *MADS55* are activated by the *G1*-Like (G1L/ALOG) transcriptional activator *G1L5* (*TAWAWA1*/*TAW1*), but activation of *MADS47* has not been reported [@Yoshida:2013ff]. Overexpression of *TAW1* increases the production of secondary branches and leads to the formation of tertiary branches, resulting in higher grain yield [@Yoshida:2013ff]. *TAW1* was recovered from the same cluster as *MADS55*, and the presence and co-expression of genes that are involved in inflorescence development along with other TF genes indicates that the clusters capture biologically meaningful groups of genes, including uncharacterized signalling components. Only three ALOG genes (which are annotated with the Pfam domain PF04852) were detected in the dataset. Two were retrieved from clusters `r getClusterName('dec', 'small')` and `r getClusterName('dec', 'dec')` (*G1L2* and *TAW1* respectively), and the third (*G1L1*) has a similar expression pattern (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`b). *taw1* missense mutants have small inflorescences with reduced grain yield, but this phenotype was stronger when *TAW1* was targeted by RNA interference [@Yoshida:2013ff]. This may suggest functional redundancy between the three co-expressed ALOG genes, *G1L1*, *G1L2* and *TAW1*, with more than one affected by the RNAi construct used against *TAW1*. Although the increase in secondary branches linked to *TAW1* overexpression suggests that it acts by suppression of SM identity, perhaps via downstream factors that may promote BM activity such as the *STMADS11*-like genes, *taw1* plants also have fewer primary branches [@Yoshida:2013ff], and the peak expression in the RM of *G1L1*, *G1L2*, *TAW1*, *MADS22* and *MADS55* suggests an additional role before the transition from apical to axillary meristem.

@@

Compared to the switch from apical to axillary meristem, more gradual changes in gene expression were apparent during the sequential transition from PBM to SM (`r I(pasteLabel("Figure ", figCount, "f_mfuzzClusters"))`). This transition corresponds to the acquisition of determinate fate by the axillary meristem. The enrichment of MADS genes in the SM sample and their overrepresentation in cluster `r getClusterName('small', 'inc')` is consistent with their involvement in flower development in rice and other plants (`r I(pasteLabel("Figure ", figCount, "f_gsea"))`). In the eudicot *A. thaliana*, *APETALA1* (*AP1*) is involved in determining FM identity. *AP1* also interacts with members of the *AGL2*-like clade (*SEPALLATA* MADS-domain factors), *APETALA3* (*AP3*) and *PISTILLATA* (*PI*), to specify petal identity, and *AGL2*-like clade members also interact with *AGAMOUS* (*AG*) to specify stamens [recently reviewed by @Prunet:2013ki]. In rice, functions similar to those provided by *A. thaliana* *PI* and *AP3* genes, related to specification of stamens and lodicules, appear to be performed redundantly by *MADS2*, *MADS4* and *SUPERWOMAN 1* (*SPW1*/*MADS16*) [@Nagasawa:2003gu; @Yao:2008dy]. Stamen and carpel identity are specified by the expression of *RICE AGAMOUS* (*RAG*/*MADS3*) and *MADS58* [@Kyozuka:2002vm; @Yamaguchi:2006br; @Dreni:2011bu]. Two recent studies in *A. thaliana* and the domesticated tomato, *Solanum lycopersicum*, used an LMD and RNAseq approach to investigate inflorescence development [@Park:2012dg; @Mantegazza:2014di]. These two datasets were used with the dataset presented in this article to compare the expression of the MIKC^C^-type subgroup of MADS genes [@Henschel:2002ui], suggesting several differences between the species (`r I(pasteLabel("Figure S", s_figCount, "sf_madsTree"))`). *A. thaliana* *AGL6* is the only gene from the *AGL6*-like clade that is more highly expressed in indeterminate meristems than determinate meristems. *agl6* mutants have fewer axillary buds on the inflorescence stem, suggesting that *AGL6* promotes axillary meristem formation [@Huang:2012de]. The rice genes from the *AGL6*-like clade, *MFO* (*MADS6*) and *MADS17*, redundantly promote floral meristem determinancy [@Ohmori:2009kj], and are correspondingly enriched in determinate meristems. Expression of the putative tomato *AGL6*-like gene *SOLYC01G093960.2* is also enriched in determinate meristems, suggesting that the role of *AGL6* in axillary meristem formation may be specific to the Arabidopsis lineage. In contrast, consistent with their functions in floral organ specification in rice and *A. thaliana*, genes in the *AGL2*-like clade are generally more highly expressed in determinate meristems.

The strong enrichment of SQUAMOSA promoter binding protein-like (SPL/SBP) genes in the PBM and their presence in cluster `r getClusterName('small', 'dec')` suggests involvement in the early stages of inflorescence development. *WEALTHY FARMER'S PANICLE* (*WFP*/*SPL14* or *IDEAL PLANT ARCHITECTURE 1*/*IPA1*), *SPL3*, *SPL12* and *SPL17* are all highly expressed in the RM and PBM, gradually decrease in expression and have weakest expression in the SM (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c), suggesting a redundant function in indeterminate axillary meristems. These genes are all putative targets of *miR156*, a microRNA that causes reduced panicle size and delayed flowering when overexpressed [@Xie:2006dc; @Wang:2015jx], and *SPL* gene expression is controlled by *miR156* and *miR529* [@Jeong:2011gv; @Wang:2015jx]. The precursor for another microRNA, *MIR319A*, was recovered from the same cluster as *SPL3* and *SPL12* and has a similar expression pattern to these genes, *SPL17* and *WFP* (`r I(pasteLabel("Figure ", figCount, "f_alogFamily"))`c). In *A. thaliana*, *miR319a* targets *TEOSINTE BRANCHED*/*CYCLOIDEA*/*PROLIFERATING CELL FACTORS* (*TCP*) transcription factors and is involved in floral organ development [@Palatnik:2003kt; @Nag:2009fc]. No direct interaction with *SPL* genes or role in reproductive development has been reported for *miR319a* in rice. Overexpression of *WFP*, which was detected by *in situ* hybridization experiments in the RM and branch meristems in wild-type plants, is associated with increased branching and grain yield [@Jiao:2010ft; @Miura:2010it]. WFP binds to the promoter region and appears to cause upregulation of the *DENSE AND ERECT PANICLE 1* (*DEP1*) gene [@Lu:2013hy], which is present in cluster `r getClusterName('inc', 'dec')` and encodes a major QTL for grain production and panicle morphology in high-yield *O. sativa* ssp. *japonica* varieties. Probable loss-of-function *dep1* mutants have larger SAMs with higher cell number, and reduced expression of a cytokinin oxidase gene, *GRAIN NUMBER 1A* (*GN1A*/*CKX2*) [@Huang:2009kf]. WFP also binds to the promoters of genes that are involved in the determination of inflorescence architecture via the cytokinin, auxin and gibberelic acid pathways, including *LONELY GUY 1* (*LOG1*), *PIN PROTEIN 1B* and *SLENDER RICE1* respectively, suggesting a role in the coordination of hormone signalling [@Lu:2013hy].

@@

This report describes changes in gene expression during the early stages of rice panicle development. Based on expression patterns of uncharacterized genes, transcription factors and genes with functions potentially related to hormone signalling, a switch between apical and axillary meristems was evident, followed by gradual changes during transitions between PBM and SM. This could be explained by the difference between the fates of the rachis meristem, which aborts after PBM differentiation [@Ikeda:2004tk], and the axillary meristems, which transition to determinate SM. The latter process may be comparable to the gradual meristem maturation during tomato inflorescence development [@Park:2012dg], rather than to the continual production of determinate floral meristems from the Arabidopsis inflorescence meristem. These results are expected to lead to mechanistic studies of the regulatory networks that establish gene expression in these tissues, as well as functional studies of the individual components, both of which will be vital for understanding the genetic factors influencing rice yield via panicle morphology.

@@

**`r I(pasteLabel("Figure ", figCount, "f_gsea"))`**. Geneset enrichment analysis (GSEA) of transcription factor and other transcriptional regulator families. Test statistics were calculated from the Log~2~-fold change between the stage of interest and all other stages for the expressed genes in each geneset. The absolute value of the test statistic indicates the magnitude of the geneset-level change, and a positive test statistic indicates enrichment and a negative statistic indicates depletion.
